---
title: "Prep SC data for deconv"
author: "Asa Bjorklund"
output: 
  html_document:
    self_contained: true
    highlight: tango
    df_print: paged
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true    
---

Steps:

Load the R objects from multiomics anlaysis and check them.

Subsample single nuclei data to X cells per celltype.

Run DEG detection and filter markers to have a good set of deconvolution genes.


```{r setup, include=FALSE}
knitr::opts_knit$set(progress=TRUE,verbose=TRUE)
knitr::opts_chunk$set(message=FALSE, warning=FALSE, result='hold',fig.width=10, fig.height = 8)
```


```{r packages}
suppressPackageStartupMessages(require(Matrix))
suppressPackageStartupMessages(require(dplyr))
suppressPackageStartupMessages(require(ggplot2))
suppressPackageStartupMessages(require(Seurat))
suppressPackageStartupMessages(require(gridExtra))
suppressPackageStartupMessages(require(pheatmap))
```


```{r}
cc = scale_color_gradientn(colors = c("grey", "yellow", "red", "black"))
small.leg <- theme(legend.text = element_text(size = 6), legend.key.width = unit(0.1,
    "cm"))
```


# Brain data

## Load data.

Remove all assays except "RNA"

```{r load}
indir = "input directory path"
outdir = "output directory path to save deconvolution results"
dir.create(outdir, showWarnings = F)

load("~/seurat_data_brain_cc_2022-12-22.RData")

seurat_multi

# keep only RNA assay to save space
# need to rename the 

seurat_multi = DietSeurat(seurat_multi, assays = "RNA", dimreducs = c("wnn_umap", "wnn_cc_umap"))
seurat_multi

g = gc(verbose = F)

```

Plot

```{r, fig.height=10}
cl.sel = "wsnn_cc_res.0.2"

seurat_multi = SetIdent(seurat_multi, value = cl.sel)

p1 = DimPlot(seurat_multi, label = T, reduction = "wnn_cc_umap", group.by = cl.sel) + NoAxes()
p2 = DimPlot(seurat_multi,  reduction = "wnn_cc_umap", group.by = "cell_flight_status") + NoAxes()
p3 = DimPlot(seurat_multi,  reduction = "wnn_cc_umap", group.by = "bio_origin") + NoAxes()
p4 = FeaturePlot(seurat_multi, features = "nFeature_RNA", reduction = "wnn_cc_umap") + cc + small.leg + NoAxes()


grid.arrange(p1,p2,p3,p4, ncol = 2)
```

```{r}
ct.names = c("predicted.celltype.l1","predicted.celltype.l2","predicted.celltype.l3","predicted.celltype.l4")

pl = DimPlot(seurat_multi,  reduction = "wnn_cc_umap", group.by = ct.names, label = T, combine = F) 
pl = lapply(pl, function(x) x + NoAxes() + NoLegend())
grid.arrange(grobs = pl, ncol = 2)
```


```{r, fig.height=4, fig.width=10}

p1 = VlnPlot(seurat_multi, features = "nFeature_ATAC", pt.size = 0) + NoLegend()
p2 = VlnPlot(seurat_multi, features = "nFeature_RNA", pt.size = 0) + NoLegend()
p3 = VlnPlot(seurat_multi, features = "percent_mt", pt.size = 0) + NoLegend()
grid.arrange(p1,p2,p3, ncol=2)

```

## Subsample

```{r}
table(as.matrix(seurat_multi[[cl.sel]]), seurat_multi$bio_origin)
table(seurat_multi$bio_origin)
```

Use clustering in column seurat_multi$wsnn_cc_res.0.3, smallest cluster has 128 cells. Subsample to 50 or less cells per sample -> maximum 250 cells per cluster.

```{r}
table(seurat_multi[[cl.sel]])

keep.cells = c()

for (sample in unique(seurat_multi$bio_origin)){
   cells = WhichCells(seurat_multi[, seurat_multi$bio_origin == sample], downsample = 50)
   keep.cells = c(keep.cells, cells)
}

data.sub = seurat_multi[,keep.cells]
table(data.sub[[cl.sel]])

table(as.matrix(data.sub[[cl.sel]]), data.sub$bio_origin)
table(data.sub$bio_origin)
```

## Marker genes

```{r}
mfile = file.path(outdir, paste0("markers_", cl.sel, ".csv"))
if (file.exists(mfile)){
  markers = read.csv(mfile)
}else{
  markers = FindAllMarkers(data.sub, only.pos = T, max.cells.per.ident = 500)
  write.csv(markers, file = mfile)
}

table(markers$cluster)

source("../../functions/overlap_phyper_v2.R")
l = split(markers$gene, markers$cluster)
o = overlap_phyper2(l,l,title = "All DEGs", remove.diag = T)

```

Check overlap to ST data and only use genes that are also expressed in ST.

```{r}

outdir2 = "directory to output results"

indir2 = "directory path where ST seurat object is located"
data.int = readRDS(file.path(indir2,"RR3-brain_ST_final_object.rds"))

st.genes = rownames(data.int@assays$RNA@counts)
length(st.genes)

rm(data.int)
g = gc(verbose = F)

nrow(markers)
markers = markers[markers$gene %in% st.genes,]
nrow(markers)
```


### Top 50 genes

Filter out top 50 genes (or less) per cluster.

```{r}



markers$pct.diff <- markers$pct.1 - markers$pct.2
markers$log.pct.diff <- log2((markers$pct.1 * 99 + 1)/(markers$pct.2 * 99 +
    1))
markers %>%
    group_by(cluster) %>%
    top_n(-150, p_val) %>%
    top_n(100, pct.diff) %>%
    top_n(50, log.pct.diff) -> top50
de.genes50 <- unique(as.character(top50$gene))
print(length(de.genes50))

l2 = split(top50$gene,top50$cluster)
o = overlap_phyper2(l2,l2, title = "Top 50 DEGs", remove.diag = T)

write.table(de.genes50, file = file.path(outdir2,"top50_degs_220609.csv"), quote = F, col.names = F, row.names = F)
```

### Top 100 genes

Filter out top 100 genes (or less) per cluster.

```{r}
markers %>%
    group_by(cluster) %>%
    top_n(-200, p_val) %>%
    top_n(150, pct.diff) %>%
    top_n(100, log.pct.diff) -> top100
de.genes100 <- unique(as.character(top100$gene))
print(length(de.genes100))

l2 = split(top100$gene,top100$cluster)
o = overlap_phyper2(l2,l2, title = "Top 100 DEGs", remove.diag = T)

write.table(de.genes100, file = file.path(outdir2,"top100_degs_220609.csv"), quote = F, col.names = F, row.names = F)
```

### Top 20 genes

Filter out top 100 genes (or less) per cluster.

```{r}
markers %>%
    group_by(cluster) %>%
    top_n(-100, p_val) %>%
    top_n(50, pct.diff) %>%
    top_n(20, log.pct.diff) -> top20
de.genes20 <- unique(as.character(top20$gene))
print(length(de.genes20))

l2 = split(top20$gene,top20$cluster)
o = overlap_phyper2(l2,l2, title = "Top 20 DEGs", remove.diag = T)

write.table(de.genes20, file = file.path(outdir2,"top20_degs_220609.csv"), quote = F, col.names = F, row.names = F)
```

### Variable genes

Select top 2000 variable genes and use also for deconv. OBS, do batch aware selection with `SelectIntegrationFeatures`.

```{r}
common.genes = intersect(st.genes, rownames(data.sub))
tmp = SplitObject(data.sub[common.genes,], split.by = "bio_origin")
var.genes = SelectIntegrationFeatures(tmp, nfeatures = 2000)

write.table(var.genes, file = file.path(outdir2,"hvg2000_220609.csv"), quote = F, col.names = F, row.names = F)

```



### Save data

Save the SC data, make sure that scale.data is empty and that data contains counts.

```{r}
library(SeuratDisk)

dim(data.sub@assays$RNA@scale.data)
data.sub@assays$RNA@data = data.sub@assays$RNA@counts



saveRDS(data.sub, file = file.path(outdir, "snRNAseq_subsampled.rds"))

SaveH5Seurat(data.sub, filename = file.path(outdir2,"snRNAseq_subsampled_220609.h5Seurat"), overwrite = T)
Convert(file.path(outdir2,"snRNAseq_subsampled_220609.h5Seurat"), dest = "h5ad", overwrite = T)

file.remove(file.path(outdir2,"snRNAseq_subsampled_220609.h5Seurat"))

```


### Session info

```{r}
sessionInfo()
```
