---
title: "Summary Stereoscope Brain data"
author: "Asa Bjorklund"
output: 
  html_document:
    self_contained: true
    highlight: tango
    df_print: paged
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true    
---

Stereoscope was run for all sections using top20, top50 or top100 DEGs per snRNAseq cluster. 


```{r setup, include=FALSE}
knitr::opts_knit$set(progress=TRUE,verbose=TRUE)
knitr::opts_chunk$set(message=FALSE, warning=FALSE, result='hold',fig.width=12, fig.height = 10)
```


```{r packages}
suppressPackageStartupMessages(require(Matrix))
suppressPackageStartupMessages(require(dplyr))
suppressPackageStartupMessages(require(ggplot2))
suppressPackageStartupMessages(require(Seurat))
suppressPackageStartupMessages(require(STutility))
suppressPackageStartupMessages(require(gridExtra))
suppressPackageStartupMessages(require(pheatmap))
```

```{r}
small.leg <- theme(legend.text = element_text(size=6), legend.key.width = unit(0.1,"cm"))
cc = scale_color_gradientn(colors = c("grey", "yellow", "red", "black"))

cl.colors = c("#e6194B", "#3cb44b", "#ffe119", "#4363d8", "#f58231", "#911eb4", "#42d4f4",
    "#f032e6", "#bfef45", "#fabed4", "#469990", "#dcbeff", "#9A6324", "#fffac8",
    "#800000", "#aaffc3", "#808000", "#ffd8b1", "#000075", "#a9a9a9", "#000000")
```


## Load ST data

Create seurat object with correct clustering etc.  Cannot load the seurat v4 object in this env, so instead saved the parts needed in `prep_st_brain_220512_for_deconv.Rmd`. 


```{r load}

indir1 = "directory path of ST seurat object"
load(file.path(indir1,"RR3-brain_ST_final_object.rds"), verbose = T)

data.int = CreateSeuratObject(counts, meta.data = meta)

data.int@tools[["Staffli"]] = staffli
data.int[["umap"]] <- CreateDimReducObject(embeddings = umap, key = "UMAP_")
     
     
```



```{r}
p = DimPlot(data.int, group.by = "seurat_clusters", label = T) + NoAxes()
print(p)

p = DimPlot(data.int, group.by = "annot", label = T, repel = T) + NoAxes() + small.leg
print(p)

p = ST.FeaturePlot(data.int, features = "nFeature_RNA", ncol = 4, show.sb = FALSE)

```




## Load Stereoscope results


```{r}
types = c("top20_degs","top50_degs","top100_degs", "hvg2000")


input_dir = "directory path where deconvolution results are saved"
samples = sort(unique(data.int$section.name))

for (type in types){
  input_dir2 = file.path(input_dir, paste0(type,"_220609"))
  print(sprintf("Reading from %s",input_dir2))
  alldata = list()
  for (sample in samples) {
    dfile = file.path(input_dir2, sample, "proportions.csv")
    tmp = read.csv(dfile, row.names = 1)
    alldata[[sample]] = tmp
  }

  all = Reduce(rbind, alldata)
  all = t(all)
  all = all[, colnames(data.int)]

  data.int[[type]] <- CreateAssayObject(counts = all)
}
  
celltypes = rownames(all)
ct.colors = cl.colors[1:length(celltypes)]
names(ct.colors)=celltypes
```

Save a seurat object with all the deconv data

```{r}
savefile = file.path(input_dir, "summary", "sobject_brain_deconv_220609.Rds")
saveRDS(data.int, file = savefile)
```


### Violin plots

Plot predicted proportion of each SN-cluster onto each ST-cluster.

```{r, fig.height=10, results='asis'}
for (type in types){
  cat("For",type, "\n")
  p = VlnPlot(data.int, features = celltypes, assay = type, pt.size = 0, ncol=5)
  print(p)
}
```

### Per section

Plot predicted proportion of each SN-cluster for each section.

```{r}
cell.prop.section = list()
samples = sort(unique(data.int$section.name))

for (type in types){
  cell.prop = list()
  for (s in samples) {
      cp = rowSums(data.int@assays[[type]]@counts[,data.int$section.name == s])
      cp = cp/sum(cp)
      cell.prop[[s]] = cp
  }
  cell.prop = Reduce(cbind, cell.prop)
  colnames(cell.prop) = samples
  cell.prop.section[[type]] = cell.prop
}


```


```{r, fig.height=8}

for (type in types){
  tot.prop = data.frame(cell.prop.section[[type]])
  tot.prop$celltype = factor(rownames(tot.prop), levels = celltypes)

  l = reshape2::melt(tot.prop, id.vars = "celltype")

  p5 = ggplot(l, aes(x = variable, y = value, fill = celltype)) + geom_bar(position = "fill",
    stat = "identity") + RotatedAxis() + scale_fill_manual(values = ct.colors) + ggtitle(type)
  
  print(p5)
}
```

### Per cluster

Plot predicted proportion of each SN-cluster for each ST-cluster

```{r}
cell.prop.cl = list()
clusters = levels(data.int$seurat_clusters)

for (type in types){
  cell.prop = list()
  for (cl in clusters) {
      cp = rowSums(data.int@assays[[type]]@counts[,data.int$seurat_clusters == cl])
      cp = cp/sum(cp)
      cell.prop[[cl]] = cp
  }
  cell.prop = Reduce(cbind, cell.prop)
  colnames(cell.prop) = clusters
  cell.prop.cl[[type]] = cell.prop
}


```


```{r, fig.height=8}

for (type in types){
  tot.prop = data.frame(cell.prop.cl[[type]])
  tot.prop$celltype = factor(rownames(tot.prop), levels = celltypes)

  l = reshape2::melt(tot.prop, id.vars = "celltype")

  p5 = ggplot(l, aes(x = variable, y = value, fill = celltype)) + geom_bar(position = "fill",
    stat = "identity") + RotatedAxis() + scale_fill_manual(values = ct.colors) + ggtitle(type)
  
  print(p5)
}
```

### Celltypes per spot

Proportion of the most abundant celltype per spot.

```{r, fig.height=10}

top.ct = sapply(types, function(x) apply(data.int@assays[[x]]@counts,2,max))

par(mfrow = c(2,2))
s = sapply(types, function(x) hist(top.ct[,x], 100, main = x))
```


Number of celltypes per spot (only counting celltypes with at least 5% abundance)

```{r, fig.height=10}
cutoff = 0.05
n.ct = sapply(types, function(x) colSums(data.int@assays[[x]]@counts>cutoff))

par(mfrow = c(2,2))
s = sapply(types, function(x) hist(n.ct[,x], 20, main = x))


```


### Onto sections

Plot the predicted celltype proportions onto ST sections


```{r, fig.height=16, results='asis'}
for (type in types){
  cat("For",type, "\n")
  data.int@active.assay = type
  p = ST.FeaturePlot(data.int, features = celltypes, ncol = 4, grid.ncol = 3, show.sb = FALSE)
  print(p)
}
```


Or to PDF for each celltype for higher resolution

```{r}
for (type in types){
  plotfile = file.path(input_dir,type,sprintf("celltype_sections_%s.pdf",type))
  data.int@active.assay = type
  pdf(plotfile, height = 7, width = 10)
  for (celltype in celltypes){
    p = ST.FeaturePlot(data.int, features = celltype, ncol = 4, show.sb = FALSE,pt.border = F)
    print(p)
  }
  dev.off()
}
```

### Onto umap

Plot celltype proportions on the ST umap.

```{r, fig.height=12}
for (type in types){
  cat("For",type, "\n")
  data.int@active.assay = type
  p = lapply(celltypes, function(x) FeaturePlot(data.int, features = x, combine = T) + cc + small.leg + NoAxes())
  grid.arrange(grobs = p, ncol=4, top = type)
}
```


### Compare methods

For each celltype, plot distribution of each celltype.

```{r, fig.height=4}

all = lapply(types, function(x) {
  d = data.frame(t(data.int@assays[[x]]@counts))
  d$type = x
  d$sample = data.int$section.name
  d$cluster = data.int$seurat_clusters
  m = reshape2::melt(d)
  return(m)
})

stats = Reduce(rbind,all)


for (ct in celltypes){
  s = stats[stats$variable == ct,]
  p = ggplot(s, aes(x=cluster,y=value, fill=type)) + geom_boxplot(outlier.size = .1) + RotatedAxis() + ggtitle(sprintf("SN cluster %s",ct))
  print(p)
}


```



### Celltype co-occurence

For now, run with only top100_deg data. Define presence of a celltype in a spot if it has a proportion > 5%. 

```{r}
source("../../functions/overlap_phyper_v2.R")

cutoff = 0.05

prop = data.int@assays$top100_degs@counts
co.occ = sapply(celltypes, function(x) which(prop[x,] > cutoff))

o = overlap_phyper2(co.occ, co.occ, title = "All", remove.diag = T)

```

Heatmap with phyper p-value for co-occurence. 

Create a graph based on the p-values from above. Only draw edges if the p-value is > 0.01 and weight edges based on -log10(pvalue).

```{r}
library(igraph)
make.igraph = function(ptest, cutoff = 0.01, title = "Celltype co-occurence", color = ct.colors) {
    pvals = ptest$P[-ncol(ptest$P), -nrow(ptest$P)]

    pseudo = min(pvals[pvals > 0 & !is.na(pvals)])
    pvals[pvals == 0] = pseudo

    # filer all edgens with pval > cutoff
    pvals[pvals > cutoff] = 1

    # convert to weights
    w = -log10(pvals)
    g = graph_from_adjacency_matrix(w, mode = "undirected", weighted = T, diag = F)
    set.seed(13)
    g.layout <- layout_with_graphopt(g)
    # g.layout<-layout_with_lgl(g, maxiter = 500)
    V(g)$color = color
    # scale point size by proportions
    nC = ptest$M[-ncol(ptest$M), ncol(ptest$M)]

    # plot.igraph(g,layout=g.layout, vertex.label.color='black',
    # edge.width=E(g)$weight/10, main = title)
    plot.igraph(g, layout = g.layout, vertex.label.color = "black", edge.width = E(g)$weight/50,
        vertex.size = nC/sum(nC) * 200, vertex.label.dist = 2, main = title)

    return(list(g = g, layout = g.layout))
}

g = make.igraph(o, title = "All")
```




### Clustering with deconv

Take the deconv matrices and redo clustering and umap based on celltype composition.

```{r}
nPC = 10

data.int$expression_clust = data.int$seurat_clusters
    p1 = ST.FeaturePlot(object = data.int, 
               features = "expression_clust", 
               pt.size = 1.5, 
               ncol = 4,
               cols = cl.colors,
               dark.theme = F)

print(p1)    
    


for (type in types){
  data.int@active.assay = type
  data.int = ScaleData(data.int)
  pc.name = paste0("pca_","type")
  umap.name = paste0("umap_","type")
  data.int = RunPCA(data.int, features = celltypes, reduction.name = pc.name, verbose = F)
  data.int = RunUMAP(data.int, reduction = pc.name, reduction.name = umap.name, dims = 1:nPC, verbose = F)
  data.int = FindNeighbors(data.int, reduction = pc.name, dims = 1:nPC, verbose = F)
  data.int = FindClusters(data.int, verbose = F)
 
  cl.name = paste0(type,"_snn_res.0.8")
  p2 = DimPlot(data.int, group.by = cl.name, label = T, cols = cl.colors) + NoAxes() + ggtitle(type)
  print(p2)
  p3 = ST.FeaturePlot(object = data.int, 
               features = cl.name, 
               pt.size = 1.5, 
               ncol = 4,
               cols = cl.colors,
               dark.theme = F) + ggtitle(type)
  print(p3)
  
  
}



```



<details>
  <summary>**Session Info**</summary>

```{r}
sessionInfo()
```
