---
title: "Comparison of DEGs with RR3 bulk RNAseq data, hypergeometric test for gene overlap"
author: "Yuvarani Masarapu"
date: "2023-10-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read the 825 DEGs from the multiomics clusters
```{r}
library(readxl)
degs <- read_excel("Supplementary_Table_3.xlsx", 
    skip = 1) # pdf version available with manuscript as Supplementary Table 3
```

Add expression labels 'upregulated' and 'downregulated' based on positive or negative logFC values for each gene
```{r}
degs$logFC <- as.numeric(degs$logFC)

degs$expression_label <- ifelse(degs$logFC>=0, "upregulated", "downregulated")
```

# Read the RR3 bulk data DEGs list
```{r}
library(readr)
RR3_degs <- read_csv("/RR3_brain_bulk_GLDS352/GLDS-352_rna_seq_differential_expression.csv")

RR3_degs <- RR3_degs[RR3_degs$`P.value_(Space Flight)v(Ground Control)`<0.05, ] #629 genes with P-value less than 0.05

RR3_degs$expression_lbl <- ifelse(RR3_degs$`Log2fc_(Space Flight)v(Ground Control)`>=0, "upregulated", "downregulated")
```

# Compare to find overlaps
```{r}
# Total 11 genes that overlap
overlap_genes <- intersect(degs$Gene, RR3_degs$SYMBOL)

d1 <- degs[(degs$Gene %in% overlap_genes), c(1,7)]
colnames(d1) <- c("gene", "label")

r1 <- RR3_degs[(RR3_degs$SYMBOL %in% overlap_genes), c(2,29)]
colnames(r1) <- c("gene", "label")

require(sqldf)

a1Ina2 <- sqldf('SELECT * FROM d1 INTERSECT SELECT * FROM r1') # 2 genes that have the same directional change in expression as compared to the RR3 ST and multiomics datasets in this study.
```

# Hypergeometric distribution test 

* We test the overlap of the 11 genes found above using phyper()
* Null hypothesis: The overlap of 11 genes from geneset1 (multiomics DEGs) and geneset2 (RR3 bulkRNAseq DEGs) is a random sampling effect.

## Pull names of all mouse genes captured by the multiomics RR3 brain dataset
```{r}
mo_brain <- readRDS("mo_brain_2022-08-26_with_annotations.rds")

DefaultAssay(mo_brain) <- "RNA"
all_genes <- rownames(mo_brain) #32170 Mus musculus genes from multiomics dataset
```

## Pull the genes captured by the bulkRNAseq data
```{r}
library(readr)
bulkRNAseq_counts <- read_csv("GLDS-352_rna_seq_RSEM_Unnormalized_Counts.csv")
#BiocManager::install("org.Mm.eg.db")
library(org.Mm.eg.db)

ensm.ids <- bulkRNAseq_counts$...1 #56748 Mus musculus genes from bulkRNAseq dataset
bulkRNA_genes <- mapIds(org.Mm.eg.db, ensm.ids, keytype="ENSEMBL", column="SYMBOL", multiVals = "first") #convert ensemble ids into gene symbols
```

```{r}
all_genes <- c(all_genes, bulkRNA_genes) #88918 genes in total that is tested 
geneset1 <- degs$Gene
geneset2 <- RR3_degs$SYMBOL
commongenes <- intersect(geneset1, geneset2)
```

## phyper() run here
```{r}
hyper_pval <- phyper(
  q = length(commongenes), 
  m = length(geneset1),
  n = length(all_genes) - length(geneset1),
  k = length(geneset2), lower.tail=FALSE
)

hyper_pval
# [1] 0.01582549
```
