---
title: "L_R Brain"
author: "Christer Sylvén"
date: "`r format(Sys.time(), '%y-%m-%d')`"
output: 
  html_document:
    self_contained: true
    highlight: tango
    df_print: paged
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true 
---

```{r setup, include=FALSE}
knitr::opts_knit$set(progress=TRUE,verbose=TRUE)
knitr::opts_chunk$set(message=FALSE, warning=FALSE, result='hold',fig.width=12, fig.height = 10)
```

```{r}
suppressPackageStartupMessages(require(data.table))
suppressPackageStartupMessages(require(dplyr))
suppressPackageStartupMessages(require(biomaRt))
suppressPackageStartupMessages(require(Rsamtools))
suppressPackageStartupMessages(require(Signac))
suppressPackageStartupMessages(require(ggplot2))
suppressPackageStartupMessages(require(Seurat))
suppressPackageStartupMessages(require(openxlsx))
suppressPackageStartupMessages(require(reshape2))
suppressPackageStartupMessages(require(readxl))
```


```{r}
path1 <- 'path to directory where the multiomics data resides'


file <- 'multiomics data object in the format .RData'
load_object <- paste0(path1, file)
load_object

load(load_object)

seurat_multi
names(seurat_multi[[]])

DefaultAssay(seurat_multi) <- "SCT"

Idents(seurat_multi) <- seurat_multi$wsnn_cc_res.0.2
table(Idents(seurat_multi))

brain_LR <- subset(seurat_multi, idents = c('4', '5', '11', '14'))

brain_LR_counts <- as.data.frame(brain_LR@assays$SCT@counts)

brain_LR_metadata <- data.frame(Cell = rownames(brain_LR@meta.data),cell_type = brain_LR@meta.data$wsnn_cc_res.0.2)

brain_LR_metadata[,2] <- factor(brain_LR_metadata[,2] )

dim(brain_LR_counts)
dim(brain_LR_metadata)

rm(seurat_multi)

```


```{r}
x <- rownames(brain_LR_counts)
head(x)
length(x)

human = useMart("ensembl", dataset = "hsapiens_gene_ensembl", verbose = TRUE, host = "dec2021.archive.ensembl.org")
mouse = useMart("ensembl", dataset = "mmusculus_gene_ensembl", verbose = TRUE, host = "dec2021.archive.ensembl.org")



genesV2 = getLDS(attributes = c("mgi_symbol"), 
                 filters = "mgi_symbol", 
                 values = x, 
                 mart = mouse, 
                 attributesL = c("hgnc_symbol"), 
                 martL = human, 
                 uniqueRows=T)
head(genesV2)
```

Mouse genes with human ortologs

```{r}
# Generally  mouse human ortologs
dim(genesV2) 
genesV2 <- distinct(genesV2, MGI.symbol, .keep_all = TRUE)
dim(genesV2) # Unique mouse humna ortologs
xx <- x[x %in% genesV2[,1]]
length(xx) # unique mouse humna ortologs in space data

genesV3 <- genesV2
genesV3 <- genesV3[match(xx, genesV3[,1]),] # correct order of orthologs
head(genesV3)
brain_LR_counts <- brain_LR_counts[xx,]
dim(brain_LR_counts)

# Human gene symbols needed for CellphoneDB
row.names(brain_LR_counts) <- make.names(genesV3[,2], unique=TRUE)
head(brain_LR_counts)
```


```{r}

# normalize counts

brain_LR_counts <- NormalizeData(object = brain_LR_counts)


setwd(path1)
write.table(brain_LR_counts, 'brain_LR_counts.txt', sep='\t',  col.names = T, quote=F)


brain_LR_metadata$cell_type[which(brain_LR_metadata$cell_type == '4')] <- '4 Astrocytes'
brain_LR_metadata$cell_type[which(brain_LR_metadata$cell_type == '5')] <- '5 Gluta Telencephalon'
brain_LR_metadata$cell_type[which(brain_LR_metadata$cell_type == '11')] <-'11 Gaba Telencephalon'
brain_LR_metadata$cell_type[which(brain_LR_metadata$cell_type == '14')] <-'14 Hypothalamus'
head(brain_LR_metadata)

write.table(brain_LR_metadata,
            file = 'cellPhone/brain_LR_metadata.txt',
            quote = F,
            col.names = T,
            row.names = F,
            sep = '\t')

```

```{r}
DE_genes_SC <- data.frame() # Create empty data.frame
  excel_sheet <- read.xlsx(paste0(path1,"formatted_de_genes_2022_06-09.xlsx"), colNames  = TRUE, sheet = 2) # Read sheet i
  DE_genes_SC <- rbind(DE_genes_SC, excel_sheet) 
  Cl4 <-  DE_genes_SC[1:93,1:4] 
  colnames(Cl4) <-  Cl4[1,]
  Cl4 <- Cl4[-1,]
 Cl4$cluster <- '4'
 Cl4 <- Cl4[,c(5,1:4)]
 
   Cl5 <-  DE_genes_SC[1:171,5:8] 
  colnames(Cl5) <-  Cl5[1,]
  Cl5 <- Cl5[-1,]
   Cl5$cluster <- '5'
 Cl5 <- Cl5[,c(5,1:4)]
  
   Cl11 <-  DE_genes_SC[1:47,17:20] 
  colnames(Cl11) <-  Cl11[1,]
  Cl11 <- Cl11[-1,]
   Cl11$cluster <- '11'
 Cl11 <- Cl11[,c(5,1:4)]
  
    Cl14 <-  DE_genes_SC[1:28,21:24] 
  colnames(Cl14) <-  Cl14[1,]
  Cl14 <- Cl14[-1,]
   Cl14$cluster <- '14'
 Cl14 <- Cl14[,c(5,1:4)]
 
   brain <- rbind(Cl4, Cl5, Cl11, Cl14)
```


```{r}
x <- brain[,2]
head(x)
length(x)

human = useMart("ensembl", dataset = "hsapiens_gene_ensembl", verbose = TRUE, host = "dec2021.archive.ensembl.org")
mouse = useMart("ensembl", dataset = "mmusculus_gene_ensembl", verbose = TRUE, host = "dec2021.archive.ensembl.org")

genesV2 = getLDS(attributes = c("mgi_symbol"), 
                 filters = "mgi_symbol", 
                 values = x, 
                 mart = mouse, 
                 attributesL = c("hgnc_symbol"), 
                 martL = human, 
                 uniqueRows=T)
head(genesV2)
```

# Mouse genes with human ortologs for formatted DE genes

```{r}
# Generally  mouse human ortologs
dim(genesV2) 
genesV2 <- distinct(genesV2, MGI.symbol, .keep_all = TRUE)
dim(genesV2) # Unique mouse humna ortologs
xx <- x[x %in% genesV2[,1]]
length(xx) # unique mouse humna ortologs in space data

genesV3 <- genesV2
genesV3 <- genesV3[match(xx, genesV3[,1]),] # correct order of ortologs
genesV4 <- brain
head(genesV3)

brain2 <- brain[brain$Gene %in% genesV2[,1],]
brain2$Gene <- genesV3[,2]
head(brain2)

```
  
 


```{r}

brain2$cluster[which(brain2$cluster == '4')] <- '4 Astrocytes'
brain2$cluster[which(brain2$cluster == '5')] <- '5 Gluta Telencephalon'
brain2$cluster[which(brain2$cluster == '11')] <-'11 Gaba Telencephalon'
brain2$cluster[which(brain2$cluster == '14')] <-'14 Hypothalamus'

brain_up <- brain2[which(brain2$logFC>0),]
  brain_down <- brain2[which(brain2$logFC<=0),]
 

  
  write.table(brain_up[,1:2],
            file = 'cellPhone/brain_LR_up.txt',
            quote = F,
            col.names = T,
            row.names = F,
            sep = '\t')
  
   write.table(brain_down,
            file = 'cellPhone/brain_LR_down.txt',
            quote = F,
            col.names = T,
            row.names = F,
            sep = '\t')
```

```{r}
# install cellphonedb as a system resource and use path as below
system('python3 -m venv cpdb')
system('source cpdb/bin/activate')
system('pip install CellPhoneDB==3.0.2')
#system('cellphonedb method degs_analysis --help')

#system('python3 -m pip install --upgrade pip')

plot <- "cellphonedb method degs_analysis  /../../Space2/cellPhone/brain_LR_metadata.txt  /../../Space2/cellPhone/brain_LR_counts.txt  /../../Space2/cellPhone/brain_LR_up.txt  --counts-data=gene_name --output-path=/../../Space2/cellPhone/out/"


plot <- "cellphonedb plot dot_plot --/../../Space2/cellPhone/brain_LR_metadata.txt --pvalues-path=/../../Space2/cellPhone/out/pvalues.txt
--means-path=/../../Space2/cellPhone/out/means.txt
--output-path=/../../Space2/cellPhone/out/  --output-name Brain_up.pdf"

address = "/Library/Frameworks/Python.framework/Versions/3.8/bin/"

#plot <- "cellphonedb plot heatmap_plot metadata_M7.txt"
plot_address <- paste0(address, plot)
plot_address
system(plot_address)

```

But shortly, we recommend to do differential expression for the same cluster/cell type between your two groups. This will tell you what is significantly upregulated in each of your group.
Then, you can give this list of differentially expressed genes (DEGs) to cellphoneDB to extract the interactions where these DEGs are participating.
 
You can run:
cellphonedb method degs_analysis your_meta.txt your_counts.txt your_DEGs_upregulated_in_groupA.txt
cellphonedb method degs_analysis your_meta.txt your_counts.txt your_DEGs_upregulated_in_groupB.txt
 
If you are using Seurat, you might want to follow https://github.com/ventolab/CellphoneDB/blob/master/notebooks/0_prepare_your_data_from_Seurat.ipynb
 
Hope this is of help,
 
Best
 
Luz

# DotPlot for relevant interactions flight upregulated ligand_recceptor pairs

```{r}
brain_celltype_interaction <- c(
  '4 Astrocytes|5 Gluta Telencephalon',
  '4 Astrocytes|11 Gaba Telencephalon',
  '4 Astrocytes|14 Hypothalamus',
  '5 Gluta Telencephalon|4 Astrocytes',
  '5 Gluta Telencephalon|11 Gaba Telencephalon',
  '5 Gluta Telencephalon|14 Hypothalamus',
  '11 Gaba Telencephalon|4 Astrocytes',
  '11 Gaba Telencephalon|5 Gluta Telencephalon',
  '11 Gaba Telencephalon|14 Hypothalamus',
  '14 Hypothalamus|4 Astrocytes',
  '14 Hypothalamus|5 Gluta Telencephalon',
  '14 Hypothalamus|11 Gaba Telencephalon')

brain_up_LR <- c(
'CADM1_NECTIN3',
'CADM3_CADM1',
'CADM1_CADM1',
'EGFR_NRG1',
'EGFR_TGFA',
'EGFR_COPA',
'PTN_PTPRZ1',
'EFNA5_EPHA5',
'VEGFA_GRIN2B')

write.table(brain_celltype_interaction, file = '/../../SPACE2/cellPhone/in/Brain_celltype_interaction.txt', sep = '\t', quote = F, row.names = F, col.names = F)

write.table(brain_up_LR, '/../../SPACE2/cellPhone/in/brain_up_LR_rows.txt', sep = '\t', quote = F, row.names = F, col.names = F)

write.table(brain_celltype_interaction, file = '/../../in/Brain_celltype_interaction.txt', sep = '\t', quote = F, row.names = F, col.names = F)

write.table(brain_up_LR, '/../../in/brain_up_LR_rows.txt', sep = '\t', quote = F, row.names = F, col.names = F)

address = "/Library/Frameworks/Python.framework/Versions/3.8/bin/"

plot <- 'cellphonedb plot dot_plot --rows /../../SPACE2/cellPhone/in/brain_up_LR_rows.txt --columns /../../SPACE2/cellPhone/in/Brain_celltype_interaction.txt --output-name brain_up_rel_interactions.png'

plot <- 'cellphonedb plot dot_plot --rows /../../SPACE2/cellPhone/in/brain_up_LR_rows.txt --columns /../../SPACE2/cellPhone/in/Brain_celltype_interaction.txt --output-name brain_up_rel_interactions.png'

plot <- 'cellphonedb plot dot_plot --rows in/brain_up_LR_rows.txt --columns in/Brain_celltype_interaction.txt --output-name brain_up_rel_interactions.png'


plot_address <- paste0(address, plot)
plot_address
system(plot_address)

```

```{r}
#subgroup output
means <-  read.delim("/../../SPACE2/cellPhone/out/means.txt", check.names = FALSE)
pvals <- read.delim("/../../SPACE2/cellPhone/out/pvalues.txt", check.names = FALSE)
#pval[pval==0] = 0.0009
dim(means)
dim(pvals)

# delete duplicates of L_R / R_L interactoin pairs
pvals0 <- pvals[!duplicated(pvals[,2]),]
#pvals1 <- pvals[unique(pvals[,1]),]

means0 <- means[!duplicated(means[,2]),]
dim(pvals0)
dim(means0)
#add L_R / R_L interaction pairs as rowname
rownames(pvals0) <- pvals0[,2]
rownames(means0) <- means0[,2]
#pvals0[1:2,1:10]
#means0[1:2,1:10]
# make dataframe w L_R R_L interaction pairs, secreted, mol 1 rec, mol 2 rec, integrin
pvals_LR <- pvals0[,c(7:9,11)]
pvals_LR <- cbind(rownames(pvals_LR), pvals_LR)
colnames(pvals_LR)[1] <- 'L_R0'

brain_celltype_interaction <- c(
  '4 Astrocytes|5 Gluta Telencephalon',
  '4 Astrocytes|11 Gaba Telencephalon',
  '4 Astrocytes|14 Hypothalamus',
  '5 Gluta Telencephalon|4 Astrocytes',
  '5 Gluta Telencephalon|11 Gaba Telencephalon',
  '5 Gluta Telencephalon|14 Hypothalamus',
  '11 Gaba Telencephalon|4 Astrocytes',
  '11 Gaba Telencephalon|5 Gluta Telencephalon',
  '11 Gaba Telencephalon|14 Hypothalamus',
  '14 Hypothalamus|4 Astrocytes',
  '14 Hypothalamus|5 Gluta Telencephalon',
  '14 Hypothalamus|11 Gaba Telencephalon')


pvals2 <- pvals0[,brain_celltype_interaction]
# to minimze number of interactions, filtered to those with p<0.01
pvals3 <- pvals2[brain_up_LR, ]
means3 <- means0[rownames(pvals3),brain_celltype_interaction]

dim(pvals3)
dim(means3)


write.table(rownames(pvals3), '/../../Space2/cellPhone/in/brain_up2_rows.txt', sep = '\t', quote = F, row.names = F, col.names = F)
plot <- 'cellphonedb plot dot_plot --rows in/brain_up2.txt --columns in/brain_celltype-interaction --output-name Brain_up.png'

```
```{r}
dotplot_LR <- function(pvals4b, means4b, namn = c(1:dim(pvals4b)[1]))
{
#pvals4b <- as.data.frame(pvals4b)
#means4b <- as.data.frame(means4b)
#rownames(pvals4b) <- pvals4b[,18]
#pvals4b <- pvals4b[,c(18, 1:12)]
pvals4b[pvals4b==0] = 0.0009

#means4b <- means4b[,c(18, 1:12)]
#means4b[means4b==0] = 1
#pvals4b <- pvals4b[namn,]
#means4b <- means4b[namn,]

#pvals4b$L_R <- rownames(pvals4b)
#means4b$L_R <- rownames(means4b)
names(pvals4b)
pvals4b <- pvals4b[,c(18, 1:12)]
means4b <- means4b[,c(18, 1:12)]

pvals5 <- reshape2::melt(pvals4b, id=c('L_R'))
means5 <- reshape2::melt(means4b, id = c('L_R'))



level1 <- pvals5[1:dim(pvals4b)[1],1]
plot.data = cbind(pvals5,log2(means5[,3]))

colnames(plot.data) = c('pair', 'clusters', 'pvalue', 'mean')

my_palette <- colorRampPalette(c("black", "blue", "yellow", "red"), alpha=TRUE)(n=399)
#my_palette <- colorRampPalette(c("blue", "red"), alpha=TRUE)(n=399)
library(ggplot2)
ggplot2::ggplot(plot.data,aes(x=clusters,y=pair)) +
  geom_point(aes(size=-log10(pvalue),color=mean)) +
  scale_color_gradientn('Log2 mean (Molecule1_Molecule2)', colors=my_palette) +
  scale_y_discrete(limits = level1,
                   labels = level1)+
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        axis.text=element_text(size=12, colour = "black"),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust=1),
        axis.text.y = element_text(size=12, colour = "black"),
        axis.title=element_blank(),
        panel.border = element_rect(size = 0.7, linetype = "solid", colour = "black"),
        legend.position="top",
        plot.margin = unit(c(0.1,1,0,2), "cm")) +
        scale_x_discrete(
          labels = c(
  '4 Astrocytes|5 Gluta Telencephalon',
  '4 Astrocytes|11 Gaba Telencephalon',
  '4 Astrocytes|14 Hypothalamus',
  '5 Gluta Telencephalon|4 Astrocytes',
  '5 Gluta Telencephalon|11 Gaba Telencephalon',
  '5 Gluta Telencephalon|14 Hypothalamus',
  '11 Gaba Telencephalon|4 Astrocytes',
  '11 Gaba Telencephalon|5 Gluta Telencephalon',
  '11 Gaba Telencephalon|14 Hypothalamus',
  '14 Hypothalamus|4 Astrocytes',
  '14 Hypothalamus|5 Gluta Telencephalon',
  '14 Hypothalamus|11 Gaba Telencephalon')) +
coord_flip()
}

dotplot_LR(pvals3, means3)
```
```{r}
# in order to edit so that all pairs are annotated as L_R and not R_L
# produce xlsx files as below and edit. R_L becomes L_R in the CONDUIT_SAN_TOP and vice versa. For means the ratio has to be changed by 1/value. Edited xlsx files are saved with suffix c and then opened to make the ggplot manually.

pvals4_calc <- function(name, LRmrna, LRcells )
{
  #pvals2 <- pvals0[,LRcells] #pval0 duplicates eliminated above
  #pvals3 <- pvals2[apply(pvals2, 1, function(x) any(-log10(x) > 2)), ]
  #pvals3 <- pvals3[LRmrna,]
 # means3 <- means0[rownames(pvals3), LRcells]
  pvals_LR_rows <- pvals_LR[rownames(pvals3),]
  pvals4 <- cbind(pvals3,pvals_LR_rows)
  means4 <- cbind(means3,pvals_LR_rows)
  writexl::write_xlsx(pvals4, paste0("/../../SPACE2/cellPhone/pvals4_d_", name, ".xlsx"))
 writexl::write_xlsx(means4, paste0("/../../SPACE2/cellPhone/means4_d_", name, ".xlsx"))
}


library(openxlsx)
library(readxl)
library(xlsx)
#install.packages(c("readxl","writexl")) 
library(readxl)
library(writexl)



pvals4_calc('Brain_up', brain_up_LR, brain_celltype_interaction)

# cellphoneDB dotplot with all pairs as L_R
#View(pvals3)
#View(means3)

# edit in excel. Change RL to LR into L_R column. Change values accordingly. Save as b for both pvals and meansfiles.
# Correct mean molecule1/molecule2 in excel with  to molecule2/molecule1
#corrected files with all pairs as L_R saved as b
pvals4b <- read_xlsx('/../../SPACE2/cellPhone/pvals4_d_brain_upb.xlsx')
means4b <- read_xlsx('/../../SPACE2/cellPhone/means4_d_brain_upb.xlsx')

dotplot_LR(pvals4b, means4b)
```
 



